<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{How to parse gatingML into a GatingSet}
-->

---
title: "How to parse gatingML into a GatingSet"
author: "Mike Jiang"
output: 
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "markup", message = FALSE, warning = FALSE)
```


This vignette demonstrates how the gatingML files exported from Cytobank can be imported into R as a GatingSet object.

```{r}
library(flowWorkspace)
fcsFiles <- list.files(pattern = "CytoTrol", system.file("extdata", package = "flowWorkspaceData"), full.names = TRUE)
xmlfile <- system.file("extdata/cytotrol_tcell_cytobank.xml", package = "flowWorkspace")
```

## Use `parse.gatingML`
The entire parsing work can be done with single convevient function `parse.gatingML`:
```{r, eval=FALSE}
gs <- parse.gatingML(xmlfile, fcsFiles)
```

## Load GatingML and FCS separately
Or you can divide the parse into several steps to have more controls.

### Load the gatingML file as a **graphGML** object
```{r}
g <- read.gatingML.cytobank(xmlfile)
class(g)
g
```
**graphGML** stores the gating hierarchy, which can be inspected by various accessors.
```{r}
getNodes(g)
getParent(g, "GateSet_722318")
getChildren(g, "GateSet_722318")

```

And the population tree can be plotted
```{r fig.width=4,fig.height=4}
plot(g)
```

The node with **dotted** border means the `tailored` gates(or sample-specific gates) are defined for that population.

### Read raw FCS files
```{r}
fs <- read.ncdfFlowSet(fcsFiles)
```

### Construct the **GatingSet**
```{r}
gs <- GatingSet(fs)
```

### Compensate the `GatingSet`
```{r}
gs <- compensate(gs, g)
```

### Extract the transformation
```{r}
trans <- getTransformations(g)
trans
```

### Transform the `GatingSet`
```{r}
gs <- transform(gs, trans)
```


### Visualize the outcome of compensation and transformation
```{r fig.width=4,fig.height=4, fig.show='hold'}
require(ggcyto)
ggcyto(gs, aes(x = CD4), subset = "root") + geom_density()
ggcyto(gs, aes(x = CD4, y = CD8), subset = "root") + geom_hex()
```


### Apply the gates to **GatingSet**
```{r}
gating(g, gs)
```


### Plot the gates
```{r}
autoplot(gs[[1]])
```

### Extract the population statistics
```{r}
getPopStats(gs, statType = "count")
```


